\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2024-01-08 paper/main.tex
moved ethics statement
edited position of fig
removed ethics statement
edited uniform prior explanation

\clearpage
\section{Experiment 2: EEG}
\subsection{Methods}
This experiment was conducted with the approval of the ethics committee of the University of Potsdam, Germany.

\subsubsection{Participants}
146 participants from the University of Potsdam participant pool took part in the experiment. Three participants were excluded because they did not fulfill the demographic requirements (bilinguals or medical history). Four participants were excluded because they finished only one out of two experimental sessions. Seven participants were excluded due to EEG artifacts (below 20 artifact-free trials in at least one condition). Additionally, 29 participants were excluded because they showed poor comprehension question accuracy (below 70\,\%) in one of the experimental sessions. The data of 103 participants (mean age: 23.5 years old, age range: 18 -- 38 years old, 81 female, 22 male) were used for the analyses presented here. These final participants were all right-handed, mono-lingual native speakers of German with normal or corrected-to-normal vision and no reported history of psychiatric or neurological disease. All participants gave written informed consent and were compensated with 40 Euros per experimental session (80 Euros in total) or course credit.

@@

\begin{figure}[H]
    \caption{Prior and posterior predicted effect sizes for syntactic and semantic interference and their interaction using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. The data with 95 \% credible intervals are shown in red.}
\begin{subfigure}[t]{\textwidth}
\caption{On the millisecond scale, i.e., for reading times.}\label{fig:model_predictions_posterior}
    \centering
    \includegraphics[width=\textwidth]{images/priorpred_vs_postpred_vs_data.png}
    \end{subfigure}

@@

The \citet{Lewis2005} model is generally used to predict reading times. The scaling factor $F$ links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R}. But this scaling factor can be adjusted to link activation to brain responses to use the model to generate predictions on the microvolt scale for ERP data. The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a uniform prior distribution ($Uniform(0.005, 0.04)$). This prior reflects the a priori belief that effects will range from 0.5 to 4 $\mu V$ with equal probability for all effect sizes in this range. This uninformative prior was chosen because of limited prior knowledge about interference effects during subject-verb dependency resolution in ERPs \citep[but see e.g.,][]{lee_garnsey, vasishth_drenhaus_2011}. We are modeling N400 amplitude differences, so larger predicted effect sizes correspond to a more negative N400. To reflect this, the predicted effects were multiplied with $-1$. As in the reading time modeling, to avoid overfitting we fitted $F$ only to the semantic interference effect in order to generate posterior predictive distributions of the syntactic and semantic interference effects. For the ERP modeling, 473 out of 20,000 sampled values of $F$ lead to simulated effects within the range of the observed effect and were thus used for posterior predictions. The prior and posterior predictions for the effects of interest on the microvolt scale in comparison to our ERP data are shown in Figure \ref{fig:model_predictions_ERP}. The prior and posterior predictions of the model overlap with the effects of interest in the data. The general effect pattern (no syntactic interference effect, but a semantic interference effect) is similar in both the prior and posterior predictive distributions. Therefore, the assumption that the relevant syntactic cue is \{$\pm$ subject-in-same-clause\} cue, rather than the \{$\pm$ grammatical subject\} cue originally assumed in previous modeling work, seems reasonable for the present ERP study as well.

>>>>

\clearpage
\section{Experiment 2: EEG}
\subsection{Methods}
\subsubsection{Participants}
146 participants from the University of Potsdam participant pool took part in the experiment. Three participants were excluded because they did not fulfill the demographic requirements (bilinguals or medical history). Four participants were excluded because they finished only one out of two experimental sessions. Seven participants were excluded due to EEG artifacts (below 20 artifact-free trials in at least one condition). Additionally, 29 participants were excluded because they showed poor comprehension question accuracy (below 70\,\%) in one of the experimental sessions. The data of 103 participants (mean age: 23.5 years old, age range: 18 -- 38 years old, 81 female, 22 male) were used for the analyses presented here. These final participants were all right-handed, mono-lingual native speakers of German with normal or corrected-to-normal vision and no reported history of psychiatric or neurological disease. All participants gave written informed consent and were compensated with 40 Euros per experimental session (80 Euros in total) or course credit.

@@

\begin{figure}[!htbp]
    \caption{Prior and posterior predicted effect sizes for syntactic and semantic interference and their interaction using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. The data with 95 \% credible intervals are shown in red.}
\begin{subfigure}[t]{\textwidth}
\caption{On the millisecond scale, i.e., for reading times.}\label{fig:model_predictions_posterior}
    \centering
    \includegraphics[width=\textwidth]{images/priorpred_vs_postpred_vs_data.png}
    \end{subfigure}

@@

The \citet{Lewis2005} model is generally used to predict reading times. The scaling factor $F$ links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R}. But this scaling factor can be adjusted to link activation to brain responses to use the model to generate predictions on the microvolt scale for ERP data. The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a uniform prior distribution ($Uniform(0.005, 0.04)$). This prior reflects the a priori belief that effects will range from 0.5 to 4 $\mu V$ with equal likelihood for all effect sizes in this range. This uninformative prior was chosen because there exists no previously published ERP data on the Van Dyke (2007) design. We are modeling N400 amplitude differences, so larger predicted effect sizes correspond to a more negative N400. To reflect this, the predicted effects were multiplied with $-1$. As in the reading time modeling, to avoid overfitting we fitted $F$ only to the semantic interference effect in order to generate posterior predictive distributions of the syntactic and semantic interference effects. For the ERP modeling, 473 out of 20,000 sampled values of $F$ lead to simulated effects within the range of the observed effect and were thus used for posterior predictions. The prior and posterior predictions for the effects of interest on the microvolt scale in comparison to our ERP data are shown in Figure \ref{fig:model_predictions_ERP}. The prior and posterior predictions of the model overlap with the effects of interest in the data. The general effect pattern (no syntactic interference effect, but a semantic interference effect) is similar in both the prior and posterior predictive distributions. Therefore, the assumption that the relevant syntactic cue is \{$\pm$ subject-in-same-clause\} cue, rather than the \{$\pm$ grammatical subject\} cue originally assumed in previous modeling work, seems reasonable for the present ERP study as well.
